---
title: "Individual Learner"
subtitle: "US Domestic Flight Delays"
author: "Ian Gledhill"
date: "12/06/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r main, code = source("Main.R"), include=FALSE, echo=FALSE}
```

```{r}
library(knitr)
```

\newpage
# Introduction

## Overview

This project will look at the delayed flight data recorded in the US for the 3 months from January to March 2018 and use data analysis to create a model to predict whether a flight would be delayed from its scheduled departure time. The data is provided by the US Department of Transport and covers `r length(levels(flights$OP_UNIQUE_CARRIER))` carriers operating from `r length(levels(flights$ORIGIN))` airports.

## Dataset

### Source

The data is provided by:

  https://www.transtats.bts.gov

  BUREAU OF TRANSPORTATION STATISTICS
  
  U.S. Department of Transportation 
  
  1200 New Jersey Avenue, SE 
  
  Washington, DC 20590 
  
  855-368-4200 


The page to download the data used in this project can be found here: 

https://www.transtats.bts.gov/Fields.asp?Table_ID=236


### Sample Size for Analysis
The DoT record more than 500,000 domestic flights per month. The period covered by this project contains `r format_number(nrow(flights))` data points. Each data point has `r length(flights)` measurements.

The original data has `r format_number(original_data_size)` observations. For the purposes of expediency a sample of `r format_number(sample_data_size)` has been randomly selected from the original data. In addition only the top `r nrow(top20Airports)` airports based on the number of departures over the data period have been analzed. These can be seen in Appendix 1.

### Shape of the Data
The data that the DoT make available falls into the 11 groupings shown below (the number of measurements in each group are shown in brackets). Not all the groupings are relevant to this project, which is only concerned with delayed departures; for example, the arrival performance at the destination airport is not relevant. The groupings that are considered useful are shown in bold.

* **Time Period** (6 catagorical)

* **Airline** (5 catagorial)

* **Origin** (9 catagorical)

* Destination (9 catagorical)

* **Departure Performance** (3 catagorical, 6 continuous)

* Arrival Performance (3 catagorical, 6 continuous)

* **Cancellations and Diversions** (3 catagorical)

* **Flight Summaries** (3 catagorical, 3 continuous )

* **Cause of Delay** (5 catagorical)

* **Gate Returns** (3 continuous)

* Diverted Airport Information (4 x sub-groups reprenting each diversion with 4 catagorical and 3 continuous)

### Overview of the Data

There are `r format_number(nrow(flights))` deomestic departures from US airports in the period covered by the data.
The break down of these flights by departure status is show below:

```{r Status_Frequency, echo=FALSE}
flights %>%
    ggplot(aes(x=DEP_STATUS)) + 
    ggtitle("All Flights", subtitle="Departure Status") +
    xlab("Departure Status") +
    ylab("Count") +
    geom_bar() +
    theme_minimal()
```  


* `r format_percent(mean(flights$DEP_STATUS == "CANCELLED"))` of flights were cancelled 

* `r format_percent(mean(flights$DEP_STATUS == "EARLY"))` of flights departed ahead of their scheduled time 

* `r format_percent(mean(flights$DEP_STATUS == "ONTIME"))` of flights departed at their scheduled time 

* `r format_percent(mean(flights$DEP_STATUS == "DELAYED"))` of flights departed after their scheduled time 

Note that the `r format_percent(mean(flights$DEP_STATUS == "UNKNOWN"))` percentage of flights (equal to `r nrow(flights[flights$DEP_STATUS == "UNKNOWN",])` flights) were not cancelled but have no depature time recorded. This is assumed to be errant data and is excluded from further analysis.

```{r Delay_Duration, echo=FALSE}
flights %>%
    filter(DEP_STATUS == "DELAYED") %>%
    mutate(DEP_DELAY_LIMIT = ifelse( DEP_DELAY > 180, 180, DEP_DELAY)) %>%
    ggplot(aes(x=DEP_DELAY_LIMIT)) + 
    geom_histogram(bins=30) +
    ggtitle("Flight Delays", subtitle="Delay Duration (> 3 hours shown as 3 hours)") +
    xlab("Delay (in minutes)") +
    theme_minimal()
```  

There are a small number of outliers which have exceptionally long delay times. These have been included in the 180 minutes bin to make the histogram more readable. We can see that most delays are within 30 minutes. 

```{r Delay_By_DayOfWeek, echo=FALSE}
flights %>%
    filter(ACTIVE) %>%
    group_by(DEP_TIME_BLK) %>%
    summarise(
      n = n(),
      n_delayed = sum(DEP_DELAY > 0),
      delay_rate = n_delayed / n
    ) %>%
    mutate(DEP_TIME_BLK_SHORT = substr(DEP_TIME_BLK,1,2)) %>%
    ggplot(aes(x=DEP_TIME_BLK_SHORT)) + 
    geom_point(aes(y = n), color="blue") +
    geom_point(aes(y=n_delayed), color="red") +
    geom_point(aes(y=delay_rate, size=delay_rate, color=delay_rate)) +
    ggtitle("Delayed Flights", subtitle="Distribution by time slot") +
    labs(col="Delay Rate", size="Delay Rate") +
    xlab("Time Slot") +
    ylab("Count") +
    theme_minimal()
```

This shows that the delay rate changes significantly depending on the time slot peels from early afternoon to early afternoon.   

```{r Delay_By_DayOfWeek2, echo=FALSE}
flights %>%
    filter(ACTIVE) %>%
    group_by(DAY_OF_WEEK) %>%
    summarise(
      n = n(),
      n_delayed = sum(DEP_DELAY > 0),
      delay_rate = n_delayed / n
    ) %>%
    ggplot(aes(x=DAY_OF_WEEK)) + 
    geom_point(aes(y = n), color="blue") +
    geom_point(aes(y=n_delayed), color="red") +
    geom_point(aes(y=delay_rate, size=delay_rate, color=delay_rate)) +
    ggtitle("Delayed Flights", subtitle="Distribution by Day Of Week") +
    labs(col="Delay Rate", size="Delay Rate") +
    xlab("Day of Week") +
    ylab("Count") +
    theme_minimal()
```

This shows that although the number of departures peeks mid-week and dips at the weekend the delay rate does not vary significantly by day. 

```{r Airport_By_AverageDelay, echo=FALSE}
flights %>%
    filter(ACTIVE) %>%
    group_by(ORIGIN) %>%
    summarise(
      n = n(),
      n_delayed = sum(DEP_DELAY > 0),
      avg_delay = mean(DEP_DELAY_NEW),
    ) %>%
    ggplot(aes(x=reorder(ORIGIN, avg_delay), ORIGIN, y=avg_delay)) + 
    geom_point(color="blue") +
    geom_label_repel(aes(label = ifelse(avg_delay>50,as.character(ORIGIN),'')),
                    box.padding   = 0.35, 
                    point.padding = 0.5,
                    segment.color = 'grey50') +
    ggtitle("Delayed Flights", subtitle="Airport by Average Delay") +
    xlab("Average Delay") +
    ylab("Count") +
    theme_minimal()
```

Some airports have considerably longer average delay times. The full list of airport codes can be seen here https://www.transtats.bts.gov/FieldInfo.asp?Field_Desc=Origin%20Airport&Field_Type=Char&Lookup_Table=L_AIRPORT&Table_ID=236&SYS_Table_Name=T_ONTIME_REPORTING&Sys_Field_Name=ORIGIN

### Training and Test Data Partitions


## Goal

The goal of this project is to create a prediction model that can estimates the probably deviation from the scheduled departure time for any  domestic flight from any US airport covered by the DoT data.


\newpage
# Approach

## Data Cleansing

#### Cancelled Flights
The aircraft will either leave early, leave on time or be delayed. The expected departure time is given by *CRS_DEP_TIME* and the number of minutes relative to this time of the actual departure is given by *DEP_DELAY*. The number of flights cancelled is `r format_number(sum(flights$CANCELLED))`. 
As these planes don't have a departure time the *DEP_DELAY* for the flight will not be available and will therefore be given a value of N/A in the data. 

## Modelling Techniques

The data set is quite large and has many features. This makes it computationally expensive. To aleviate this Principal Component Analysis (PCA) will be used to reduce the number of features.  

CART (_classification and regression trees_) and Random Forest have been adopted to build a predictive model.

After training the models:

* the best CP (_complexity parameter_) for CART was found to be `r rpart_mdl$bestTune$cp`

* the number of nodes used for Random Forests was `r rf_mdl$ntree`

Additionally, random forest is used as a comparison.

```{r important_variables}
kable(importance(rf_mdl), caption="Feature Importance")
```



\newpage
# Results

## CART

The RMSE using CART was `r rpart_RMSE`.

```{r CART_BestFit, echo=FALSE}
  plot(rpart_mdl)
```

```{r CART_Summary, echo=FALSE}
print(rpart_mdl)
```

### Random Forests

The RMSE using Random Forests was `r rf_RMSE`.

```{r RF_Summary, echo=FALSE}
print(rf_mdl)
```

\newpage
# Conclusion

There is some predictive capability but the residual error is quite large which makes the accuracy not very high. Using a larger data set may help this.

\newpage
# Appendix 1: Top Airports

Only data from the following airports has been analyzed. These are the `r nrow(top20Airports)` busiest airports.

```{r topAirports}
kable(top20Airports, caption="Top Airports")
```



\newpage
# Bibliography and References



